---
title: "Comparing groups"
subtitle: "REFERENCE SHEET"
format: 
  pdf:
    toc: true
    toc-depth: 5
    include-in-header: 
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
          commandchars=\\\{\},
          breaklines, breaknonspaceingroup, breakanywhere
        }
execute:
  message: false
  warning: false
  #echo: false
  fig-width: 8
  fig-height: 4
---

```{r}
#| label: packages

library(tidyverse)
library(flextable)
library(ggridges)
```

```{r}
#| label: data

wq <- read_csv("https://raw.githubusercontent.com/ENVS-193DD/datasets/refs/heads/main/water-quality/wq_clean.csv") |> 
  mutate(site_name = fct_relevel(site_name,
                                 "Phelps Bridge", "East Channel", "Venoco Bridge"))
```

```{r}
#| label: plot-aesthetics

# colors to use in all figures
site_cols <- c(
  "East Channel" = "orange4",
  "Phelps Bridge" = "tomato3",
  "Venoco Bridge" = "turquoise4"
)

# setting common theme elements
theme_set(
  theme_classic() +
    theme(text = element_text(size = 12),
          plot.title.position = "plot",
          legend.position = "none") 
)

```


# 1. Math

## Mean

$$
\bar{y} = \frac{1}{n}\sum_{i = 1}^ny_i
$$

$\bar{y}$ is the sample mean, which is calculated as the sum of all observations $y_i$ divided by the number of observations ($n$).

## Median

The median is determined by finding the middle value. If $n$ is odd, the median is the middle value in the dataset. If $n$ is even, the median is the mean of the two middle values.

## Standard deviation

$$
s = \sqrt{\frac{\sum(y_i - \bar{y})^2}{n - 1}}
$$

$s$ is the sample standard deviation, which is calculated as the square root of the sum of the squared difference between an observation $y_i$ and the sample mean $\bar{y}$ divided by the number of observations $n$ minus 1.

## Variance

$$
s^2 = \frac{\sum(y_i - \bar{y})^2}{n - 1}
$$

Variance is calculated as the standard deviation squared. Note that the units of variance are the units squared (e.g. g^2^, m^2^, ppt^2^).

# 2. Water quality

## a. Calculating central tendency and spread

```{r}
#| label: wq-summary

# creating new object from water quality
salinity_summary <- wq |> 
  # group by site
  group_by(site_name) |> 
  # calculate mean, median, standard deviation, variance, number of obs
  summarize(mean_sal = mean(mean_salinity_ppt),
            median_sal = median(mean_salinity_ppt),
            sd_sal = sd(mean_salinity_ppt),
            var_sal = var(mean_salinity_ppt),
            n = length(mean_salinity_ppt)) |> 
  # ungroup data frame
  ungroup()

# use flextable to create table
flextable(salinity_summary) |> 
  # change column names in table to be readable
  set_header_labels(
    site_name = "Site",
    mean_sal = "Mean",
    median_sal = "Median",
    sd_sal = "Standard deviation",
    var_sal = "Variance",
    n = "Number of surveys"
  ) |> 
  # round any columns with numbers to one decimal point
  colformat_double(j = c("mean_sal", "median_sal", "sd_sal", "var_sal"), 
                   digits = 1) |> 
  # setting column width of table
  width(width = 1,
        unit = "in")
```

## b. Visualizing distributions

Purpose: represent the distribution of variable of interest

### Histograms

**Components**

- x-axis: _continuous_ variable of interest
- y-axis: counts
- Column height represents number of observations falling into a given bin range on the x-axis

**Pay attention to:** 

- the shape of the distribution
- the center of the distribution
- the peak(s) of the distribution

```{r}
#| label: histogram

# creating object for histogram
histogram <- ggplot(data = wq,
                    # x-axis: variable of interest
                    # fill geometries by site
                    mapping = aes(x = mean_salinity_ppt,
                                  fill = site_name)) +
  # histogram geometry
  geom_histogram(bins = 8,             # using 8 bins
                 color = "black") +    # coloring outline black
  # creating different panels for each site
  facet_wrap(~ site_name) +
  # getting rid of space between columns and axis, setting limits
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 210)) +
  # filling columns
  scale_fill_manual(values = site_cols) +
  # relabelling axes, adding title
  labs(x = "Salinity (ppt)",
       y = "Count",
       title = "Histograms")

# displaying histogram
histogram
```

### Density plots

**Components:**

- x-axis: continuous variable of interest
- y-axis: probability density
- smoothed function representing the shape of the distribution
- note: area under that smoothed function = 1

**Pay attention to:**

- the shape of the **distribution**
- the center of the distribution
- the peak(s) of the distribution

```{r}
#| label: density-plot

density <- ggplot(data = wq, 
                  mapping = aes(x = mean_salinity_ppt,
                                fill = site_name)) +
  geom_density() +
  facet_wrap(~ site_name) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = site_cols) +
  labs(x = "Salinity (ppt)",
       y = "Density",
       title = "Density plots")

density
```

**Interpretation:**

Most salinity measurements from Phelps Bridge are low. Most salinity measurements from East Channel and Venoco Bridge are below 40 ppt.

### Boxplots

**Components:**

- x-axis: categories (levels in a factor)
- y-axis: variable of interest
- middle line: median
- bottom of box: first quartile (25th percentile)
- top of box: third quartile (75th percentile)
- whiskers: 1.5 $\times$ interquartile range
- points: observation values outside 1.5 $\times$ interquartile range


```{r}
#| label: boxplots

boxplot <- ggplot(data = wq,
                  mapping = aes(x = site_name,
                                y = mean_salinity_ppt,
                                fill = site_name)) +
  geom_boxplot(width = 0.5) +
  scale_fill_manual(values = site_cols) +
  labs(x = "Site",
       y = "Salinity (ppt)",
       title = "Boxplot")

boxplot
```

### Violin charts

**Note:** violin charts are density plots that are mirrored.

**Components:**

- x-axis: categories (levels in a factor)
- y-axis: variable of interest
- geometry: jittered points

**Pay attention to:**

- wide or narrow parts of the violin
- position of wide or narrow parts along y-axis

```{r}
#| label: violin

violin <- ggplot(data = wq,
                 mapping = aes(x = site_name,
                               y = mean_salinity_ppt,
                               fill = site_name)) +
  geom_violin() +
  scale_fill_manual(values = site_cols) +
  labs(x = "Site",
       y = "Salinity (ppt)",
       title = "Violin plot")

violin
```

### Strip charts

**Components:**

- x-axis: categories (levels in a factor)
- y-axis: variable of interest
- geometry: jittered points

**Pay attention to:**

- the clusters of points, if any
- position of points along the y-axis

```{r}
#| label: strip-chart

jitter <- ggplot(data = wq,
                 mapping = aes(x = site_name,
                               y = mean_salinity_ppt,
                               color = site_name)) +
  geom_jitter(height = 0,
              alpha = 0.5,
              width = 0.2,
              shape = 21) +
  scale_color_manual(values = site_cols) +
  labs(x = "Site",
       y = "Salinity (ppt)",
       title = "Strip chart")

jitter
```

## c. Visualizing central tendency and spread

### Means

**With standard deviation:**

```{r}
#| label: means-sd

jitter +
  geom_point(data = salinity_summary,
             mapping = aes(x = site_name,
                           y = mean_sal,
                           color = site_name),
             size = 2) +
  labs(title = "Strip chart with means")
```

**With standard error:**

```{r}
#| label: means-se

jitter +
  stat_summary(geom = "pointrange",
               fun.data = mean_se,
               size = 1,
               linewidth = 1) +
  labs(title = "Strip chart with means and SE")
```

**With 95% confidence interval:**

```{r}
#| label: means-ci

jitter +
  stat_summary(geom = "pointrange",
               fun.data = mean_cl_normal,
               size = 1,
               linewidth = 1) +
  labs(title = "Strip chart with means and 95% confidence interval")
```

## d. Exciting

```{r}
ggplot(data = wq,
       mapping = aes(x = mean_salinity_ppt,
                     y = site_name,
                     fill = site_name)) +
  geom_density_ridges(alpha = 0.5,
                      scale = 0.8) +
  stat_summary(geom = "pointrange",
               fun.data = mean_cl_normal,
               size = 0.5) +
  scale_fill_manual(values = site_cols) +
  scale_y_discrete(expand = c(0.01, 0.01)) +
  labs(x = "Salinity (ppt)",
       y = "Site")
```






